The evolution of gene expression in the term placenta of viviparous mammals

### 2016-03-07 manuscript/manuscript.Rnw
clean up yinotherian discussion; cite nevo book

The placenta arose at some point before the origin of therian mammals,
a group that includes both marsupials and eutherian or “placental
mammals”. The term “placental mammals” is misleading, as extant
yinothereians (such as \textit{Ornithorhynchus anatinus}, the
platypus) possess a rudimentary placenta, as do marsupials, although
the direct apposition to maternal tissues lasts for a proportionally
shorter time during gestation
\cite{Selwood.ea2006:Trophoblastandhypoblastin,Renfree2010:ReviewMarsupialsplacentalmammals}.
Marsupials also differ from eutherian mammals in that most possess
only a choriovitelline placenta, produced from the fusion of the
chorion and yolk sac, during most of gestation. Although some
eutherian mammals may possess a choriovitelline placenta at some point
in development, the major site of nutrient exchange in placentals
occurs via the chorioallantoic placenta, produced from the fusion of
the chorion and allantois
\cite{Mossman.1987:vertebratefetalmembranes}. It is unclear whether
the first placenta was more like that of marsupials or eutherians;
however, using phylogenetic methods, the ancestral eutherian was
reconstructed as possessing a relatively invasive hemochorial
placenta, in which the maternal and fetal blood supply is separated by
one tissue layer, as in humans and mouse
\cite{Elliot.ea2009:Phylogeneticevidenceforearly,Wildman.ea2006:Evolutionofmammalianplacenta}.
\DLA{To discuss maternal/fetal split with derek}

@@

Fetal tissue was collected from the placenta of eight mammalian
species (\human [human], \bonobo [bonobo], \dog [domestic dog], \cow
[domestic cow], \elephant [African savanna elephant], \opossum [gray
short-tailed opossum], \galili [blind mole rat from Upper Galilee],
\carmeli [blind mole rat from northern Israel]). Information on
specimen ID, place and time of collection, and collector is shown for
each sample in Supplemental \cref{tab:srr_numbers}. In the case of \bonobo,
\cow, and \elephant, the placenta was collected at term upon expulsion
from the mother following delivery. Placental tissue from \human was
collected upon delivery via a cesarean section. Near-term placental
tissue from \dog was collected upon the death of the mother from a car
accident. Placental tissue from \opossum was sampled at fetal stage
32, roughly one day prior to birth. Only the fetal portion of the
placenta was sampled at the maternal-fetal interface (i.e. villous
portion for \human, \bonobo, \dog, and \elephant; cotyledon for \cow;
diffuse placenta for \opossum). All tissues were collected and stored
in RNAlater (Life Technologies, Carlsbad, CA) until RNA isolation
could be performed. Tissues were homogenized in Trizol, and total RNA
was isolated following the manufacturer's instructions. RNA was
further purified using RNeasy Midi Kit (Qiagen, Valencia, CA), in
conjunction with the RNase-Free DNase Set (Qiagen, Valencia, CA),
following the manufacturer’s protocol to remove any contaminating DNA
from the sample. Total RNA was then evaluated on a Nanodrop 1000
(Thermo Scientific, Wilmington, DE) and an Agilent Bioanalyzer 2100
(Santa Clara, CA) to confirm concentration and quality of the
isolation. RNA quality was assessed using A260/A280 and A260/A230
absorbance values.

>>>>

The placenta arose at some point before the origin of extant mammals,
a group that includes marsupials, monotremes, and eutherian or
“placental mammals”. The term “placental mammals” is misleading, as
extant yinotherians (such as \textit{Ornithorhynchus anatinus}, the
platypus) possess a rudimentary placenta, as do marsupials, although
the direct apposition to maternal tissues lasts for a proportionally
shorter time during gestation
\cite{Selwood.ea2006:Trophoblastandhypoblastin,Renfree2010:ReviewMarsupialsplacentalmammals}.
Marsupials also differ from eutherian mammals in that most possess
only a choriovitelline placenta, produced from the fusion of the
chorion and yolk sac, during most of gestation. Although some
eutherian mammals may possess a choriovitelline placenta at some point
in development, the major site of nutrient exchange in placentals
occurs via the chorioallantoic placenta, produced from the fusion of
the chorion and allantois
\cite{Mossman.1987:vertebratefetalmembranes}. It is unclear whether
the first placenta was more like that of marsupials or eutherians;
however, using phylogenetic methods, the ancestral eutherian was
reconstructed as possessing a relatively invasive hemochorial
placenta, in which the maternal and fetal blood supply is separated by
one tissue layer, as in humans and mouse
\cite{Elliot.ea2009:Phylogeneticevidenceforearly,Wildman.ea2006:Evolutionofmammalianplacenta}.
\DLA{To discuss maternal/fetal split with derek}

@@

Fetal tissue was collected from the placenta of eight mammalian
species (\human [human], \bonobo [bonobo], \dog [domestic dog], \cow
[domestic cow], \elephant [African savanna elephant], \opossum [gray
short-tailed opossum], \galili [blind mole rat from Upper Galilee],
\carmeli [blind mole rat from northern Israel]
\cite{Nevo2001:adaptiveradiationofblindsubterraneanmole}). Information
on specimen ID, place and time of collection, and collector is shown
for each sample in Supplemental \cref{tab:srr_numbers}. In the case of
\bonobo, \cow, and \elephant, the placenta was collected at term upon
expulsion from the mother following delivery. Placental tissue from
\human was collected upon delivery via a cesarean section. Near-term
placental tissue from \dog was collected upon the death of the mother
from a car accident. Placental tissue from \opossum was sampled at
fetal stage 32, roughly one day prior to birth. Only the fetal portion
of the placenta was sampled at the maternal-fetal interface (i.e.
villous portion for \human, \bonobo, \dog, and \elephant; cotyledon
for \cow; diffuse placenta for \opossum). All tissues were collected
and stored in RNAlater (Life Technologies, Carlsbad, CA) until RNA
isolation could be performed. Tissues were homogenized in Trizol, and
total RNA was isolated following the manufacturer's instructions. RNA
was further purified using RNeasy Midi Kit (Qiagen, Valencia, CA), in
conjunction with the RNase-Free DNase Set (Qiagen, Valencia, CA),
following the manufacturer’s protocol to remove any contaminating DNA
from the sample. Total RNA was then evaluated on a Nanodrop 1000
(Thermo Scientific, Wilmington, DE) and an Agilent Bioanalyzer 2100
(Santa Clara, CA) to confirm concentration and quality of the
isolation. RNA quality was assessed using A260/A280 and A260/A230
absorbance values.
